Discrete Time Markov Model

### 2025-04-29 analysis/discrete-markov.qmd
Re rendering site :fire:

## Loading packages

```{r packages}
#| code-fold: true
#| code-summary: "Check out my packages"
#| warning: false

# Packages ---------------------------------------------------------------------
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

## Loading functions

```{r functions}
#| code-fold: true
#| code-summary: "Check out my functions"
#| warning: false

# Functions --------------------------------------------------------------------
extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
  # Extracts cognitive function data for specified years from a dataset.
  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
  # Arguments:
  #   - data: The input dataset containing cognitive function data.
  #   - years: A vector of years for which data should be extracted.
  # Returns:
  #   - A dataset with ID and cognitive status columns for the specified years.

  # Create dynamic column names based on the years provided
  cogfunction_cols <- paste0("cogfunction", years)
  cogtotal_cols    <- paste0("cogtot27_imp", years)

  if(cog_total == FALSE) {
    data <- data |>
      # Select only the ID column and cognitive function columns for the specified years
      select(ID, any_of(cogfunction_cols)) |>
      mutate(across(!c(ID), ~ case_when(
        .x == 1 ~ "Normal Cognition",
        .x == 2 ~ "MCI",
        .x == 3 ~ "Dementia",
        TRUE ~ NA_character_  # To handle missing/other cases
      )))
  } else {
    data <- data |>
      # Select only the ID column and cognitive function columns for the specified years
      select(ID, any_of(cogtotal_cols)) |>
      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
        string = .x,
        pattern = "cogtot27_imp", 
        replacement = "cog_score_"))
  }

  if(impute == TRUE){
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      tidyr::fill(Status, .direction = "down") |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  }

  if(absorbing == TRUE) {
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      mutate(Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status)) |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  }

  return(data)
}

pivot_and_factorise <- function(data) {
  # Converts cognitive function data from wide to long format and factorizes key variables.
  # Pivots multiple cogfunction columns into wave/status pairs and converts categorical
  # variables (Gender, Education, status) to factors with meaningful labels.
  # Arguments:
  #   - data: Dataset containing cognitive function variables in wide format
  # Returns:
  #   - Long-format dataset with factorized variables, ordered by ID and wave
  #

  # Pivoting depression variables
  # depression <- data |>
  #   select(starts_with("Total_dep")) |>
  #   tidyr::pivot_longer(
  #     cols = everything(), names_to = "Remove", values_to = "Depression") |>
  #   select(Depression)

  # Pivoting cardio variables
  # cardio <- data |>
  #   select(starts_with("Cardio_risk")) |>
  #   tidyr::pivot_longer(
  #     cols = everything(), names_to = "Remove", values_to = "Cardio") |>
  #   select(Cardio)

  data |>
    # select(!starts_with(c("Total_dep", "Cardio_risk"))) |>
    tidyr::pivot_longer(
      cols = starts_with("cogfunction"), names_to = "wave",
      names_prefix = "cogfunction", values_to = "status") |>
    mutate(
      Gender = factor(Gender, levels = c(0, 1)),
      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
      wave = factor(wave),
      status = factor(status, 
                      levels = c("Normal Cognition", "MCI", "Dementia"),
                      labels = c(1, 2, 3))) |>
    relocate(wave, .after = ID) |>
    relocate(status, .after = wave)
    #cbind(depression, cardio)
}

create_transition_table <- function(year_to, year_from) {
  # Creates a transition frequency table between two specified years of cognitive status data.
  # Calculates row sums and total observations for transition analysis.
  # Arguments:
  #   - year_to: Target year for transitions (character or numeric)
  #   - year_from: Origin year for transitions (character or numeric)
  # Returns:
  #   - List containing: 
  #     - transition frequency table
  #     - row sums
  #     - total observations
  # 
  tbl <- table(
    table_data[[paste0("HRS_", year_to)]],
    table_data[[paste0("HRS_", year_from)]],
    dnn = c(year_to, year_from)
  )

  # Calculate row sums and total
  row_sums <- rowSums(tbl)
  total <- sum(row_sums)

  # Return as a list with both the table and summary stats
  list(
    table = tbl,
    row_sums = row_sums,
    total = total
  )
}

create_transition_dataset <- function(data, transition_results) {
  # Combines multiple transition tables into a single analysis-ready dataset.
  # Formats period labels, ensures consistent factor levels, and structures data for visualization.
  # Arguments:
  #   - data: List of year pairs to process (e.g., list(c(2016,2018)))
  #   - transition_results: List containing transition tables from create_transition_table()
  # Returns:
  #   - Tidy dataset with transition frequencies between all specified periods
  # 
  data |>
    purrr::map_dfr(~ {
      period_name <- paste(.x[2], .x[1], sep = " - ")
      tbl <- transition_results[[paste(.x[2], .x[1], sep = "-")]]$table

      as.data.frame(tbl) |>
        rename(t_minus_1 = 1, t = 2) %>%  # Positional renaming
        mutate(Period = period_name, .before = t_minus_1)
    }) |>
    mutate(
      Period = stringr::str_replace(Period, "(\\d+) - (\\d+)", "\\2 - \\1"),
      Period = factor(Period, levels = c("2016 - 2018", "2018 - 2020", "2020 - 2022")),
      t_minus_1 = factor(t_minus_1, levels = c("Normal Cognition", "MCI", "Dementia")),
      t = factor(t, levels = c("Normal Cognition", "MCI", "Dementia"))
    )
}

create_transitions <- function(data, absorbing = FALSE){
  # Reshapes data from wide to long format to track cognitive status transitions over time.
  # Calculates the next wave's cognitive status for each individual and creates a transition column.
  # Optionally treats "Dementia" as an absorbing state, meaning once an individual is classified
  # with dementia, their status cannot change in subsequent waves.
  # Arguments:
  #   - data: The dataset containing cognitive status data.
  #   - absorbing: A logical flag indicating whether "Dementia" should be treated 
  #   as an absorbing state.
  # Returns:
  #   - A dataset with transition information, including current and next wave statuses.

  # Reshape the data from wide to long format to track cognitive status over waves
  data <- data |>
    select(ID, starts_with("cogfunction")) |>
    tidyr::pivot_longer(cols = !ID,
                        names_to = "Wave",
                        values_to = "Status") |>
    mutate(Wave = as.factor(stringr::str_replace(Wave, "cogfunction", ""))) |>
    # Arrange by ID and Wave to prepare for transition calculation
    arrange(ID, Wave) |>
    group_by(ID) |>
    # Get the next wave's cognitive status for each person
    mutate(next_wave_status = lead(Status)) |>
    ungroup()

  # We can optionally specify dementia as an absorbing state
  # Once an individual is classified with dementia they cannot be classified
  # with anything else 
  if(absorbing == TRUE) {
    data <- data |>
      group_by(ID) |>
      mutate(
        Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status),
        next_wave_status = ifelse(cumany(Status == "Dementia"), "Dementia", next_wave_status),
        transition = paste(Status, next_wave_status, sep = " to ")
      ) |>
      filter(Wave %in% c(2016, 2018, 2020))
  }

  # Filter out rows where either the current or next status is missing
  data <- data |>
    group_by(ID) |>
    filter(!is.na(Status), !is.na(next_wave_status)) |>
    ungroup() |>
    # Create a new column representing the transition from one status to the next
    mutate(
      transition = paste(Status, next_wave_status, sep = " to "),
      transition = factor(
        transition, 
        levels = c("Normal Cognition to Normal Cognition", "Normal Cognition to MCI",
                   "Normal Cognition to Dementia", "MCI to Normal Cognition", "MCI to MCI",
                   "MCI to Dementia", "Dementia to Dementia") 
      ))

  return(data)
}

observed_transition_matrix <- function(data) {
  # Converts transition data into a properly formatted probability transition matrix.
  # Ensures consistent state ordering and converts proportions to matrix format suitable
  # for multi-state modeling and visualization.
  # Arguments:
  #   - data: Transition dataset from create_transitions()
  # Returns:
  #   - Square transition probability matrix with states as row/column names
  # 
  # The order I want my matrix in
  state_order <- c("Normal Cognition", "MCI", "Dementia")

  data |>
    group_by(Status, next_wave_status) |>
    summarise(freq = n(),  .groups = "drop") |>
    group_by(Status) |>
    mutate(freq = round(proportions(freq), 3)) |>
    ungroup() |>
    complete(Status, next_wave_status, fill = list(freq = 0)) |>
    tidyr::pivot_wider(names_from = next_wave_status, values_from = freq, names_sort = TRUE) |>
    tibble::column_to_rownames("Status") |>
    select(all_of(state_order)) |>
    _[state_order, ] |>
    as.matrix()
}

normalise <- function(x) {
  # Normalizes a matrix so that each row sums to 1.
  # Arguments:
  #   - x: The input matrix to normalize.
  # Returns:
  #   - A normalized matrix where each row sums to 1.
  x / rowSums(x)
}

reshape_matrix <- function(matrix) {
  # Converts a transition probability matrix into a tidy format suitable for visualization.
  # Transforms the matrix into long format with explicit factor levels for cognitive states,
  # preserving the original ordering while preparing for ggplot compatibility.
  # Arguments:
  #   - matrix: A square transition probability matrix with states as row/column names
  # Returns:
  #   - Tidy data frame with columns:
  #     * from_state: Factor indicating origin cognitive state
  #     * to_state: Factor indicating destination cognitive state (reversed for plotting)
  #     * probability: Numeric transition probability values

  matrix |>
    as.data.frame() |>
    mutate(
      from_state = factor(c(
        "Normal cognition", "MCI", "Dementia"),
        levels = c("Normal cognition", "MCI", "Dementia"))
    ) |>
    reshape2::melt(
      id.vars = "from_state", 
      variable.name = "to_state", 
      value.name = "probability") |>
    mutate(to_state = factor(to_state, levels = rev(levels(to_state))))
}

plot_transition_matrix <- function(matrix, observed = TRUE) {
  # Creates a heatmap visualization of cognitive state transition probabilities.
  # Generates either observed or estimated transition plots with consistent formatting,
  # including labeled probability values and a diverging color scale for emphasis.
  # Arguments:
  #   - matrix: Tidy transition matrix from reshape_matrix()
  #   - observed: Logical flag indicating whether data represents observed (TRUE) 
  #               or estimated (FALSE) transitions
  # Returns:
  #   - ggplot heatmap object with:
  #     * State transitions as cells
  #     * Probability values displayed numerically
  #     * Custom color scale and axis formatting

  if(observed == TRUE) {
    subtitle <- "Observed state transitions between assessment waves"
  } else {
    subtitle <- "Estimated state transitions between assessment waves"
  }

  matrix |>
    ggplot(aes(x = from_state, y = to_state, fill = probability)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(
      aes(label = format(round(probability, 3), nsmall = 3)),
      size = 4.5, 
      color = "#212427",
      fontface = "bold") +
    colorspace::scale_fill_continuous_diverging(
      palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
      limits = c(0, 1), name = "Transition \nProbability") +
    labs(title = "Transition Probabilities Across Cognitive States",
         subtitle = subtitle,
         x = "Previous State (t - 1)", 
         y = "Current State (t)") +
    theme(
      axis.text = element_text(size = 10),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      legend.position = "right",
      legend.text = element_text(size = 9),
      panel.grid = element_blank()
    )
}

tran_stack_graph <- function(data) {
  # Creates stacked bar charts visualizing cognitive state transitions across time periods.
  # Shows composition of current states by previous state, faceted by observation period.
  # Arguments:
  #   - data: Transition dataset from create_transition_dataset()
  # Returns:
  #   - ggplot object showing stacked transition proportions
  # 
  data |>
    ggplot(aes(x = t_minus_1, y = Freq, fill = t)) +
    geom_col(position = "stack", colour = "black") +
    facet_wrap(~ Period, ncol = 3) +
    labs(
      x = "Previous State (t-1)", 
      y = "Count",
      title = "Outcomes by Prior Cognitive State",
      fill = "Current State (t)"
    ) +
    ggokabeito::scale_fill_okabe_ito() +
    ggeasy::easy_move_legend("bottom")
}

tran_heat_map <- function(data) {
  # Generates heatmap visualization of transition frequencies between cognitive states.
  # Uses color intensity and labeled values to show transition patterns across time periods.
  # Arguments:
  #   - data: Transition dataset from create_transition_dataset()
  # Returns:
  #   - ggplot heatmap with state transitions as cells
  # 
  data |>
    ggplot(aes(x = t_minus_1, y = t, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), color = "black", size = 3.5) +  # Add counts
    scale_fill_gradient(low = "white", high = "steelblue") +
    facet_wrap(~ Period, ncol = 3) +  # Split by time period
    labs(
      x = "Previous State (t-1)", 
      y = "Current State (t)",
      title = "Cognitive State Transitions Between Time Periods",
      fill = "Frequency"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggeasy::easy_move_legend("bottom")
}

tidy_output <- function(fit, multiplicative = FALSE) {
  output <- broom::tidy(fit, conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))

  if(any(output$y.level == "1")) {
    output <- output |>
      mutate(y.level = case_when(
      y.level == "1" ~ "MCI - NC",
      y.level == "3" ~ "MCI - Dementia"))
  } else {
    output <- output |>
      mutate(y.level = case_when(
      y.level == "2" ~ "NC - MCI",
      y.level == "3" ~ "NC - Dementia"))
  }

  if(multiplicative == FALSE) {
    output <- output |>
      mutate(
      term = case_when(
      term == "Gender1" ~ "Being female",
      term == "Education_tri1" ~ "High school degree vs. No education",
      term == "Education_tri2" ~ "Further education vs. No education",
      term == "Total_dep_2016" ~ "Depression Scores (2016)",
      term == "Total_p" ~ "Procrastination (2020)",
      term == "status_prev2" ~ "Previous state: MCI",
      term == "status_prev3" ~ "Previous state: Dementia",
      term == "wave" ~ "Time",
      TRUE ~ term))
  } else {
    output <- output |>
      mutate(
      term = case_when(
      term == "Gender1" ~ "Being female",
      term == "Education_tri1" ~ "High school degree vs. No education",
      term == "Education_tri2" ~ "Further education vs. No education",
      term == "Depression" ~ "Depression Scores (2016)",
      term == "Total_p" ~ "Procrastination (2020)",
      term == "status_prev2" ~ "Previous state: MCI",
      term == "status_prev3" ~ "Previous state: Dementia",
      term == "wave" ~ "Time",
      term == "Gender1:wave" ~ "Gender & Time",
      term == "Age:wave" ~ "Age & Time",
      term == "Education_tri1:wave" ~ "Education (0 - 1) & Time",
      term == "Education_tri2:wave" ~ "Education (0 - 2) & Time",
      term == "Depression:wave" ~ "Depression & Time",
      term == "Total_p:wave" ~ "Procrastination & Time",
      term == "status_prev2:wave" ~ "Previous state: MCI & Time",
      term == "status_prev3:wave" ~ "Previous state: Dementia & Time",
      TRUE ~ term))
  }

  return(output)
}

tidy_predictions <- function(predictions) {
  # Restructures model prediction matrices into tidy format for visualization.
  # Converts numeric codes to factor labels and reshapes multiple prediction columns
  # into key-value pairs suitable for ggplot.
  # Arguments:
  #   - predictions: Raw prediction matrix from model output
  # Returns:
  #   - Long-format dataset with probabilities for each cognitive state
  #   
  predictions |>
    as.matrix() |>
    as_tibble() |>
    mutate(
      Gender = factor(ifelse(Gender == 0, "Male", "Female"), levels = c("Male", "Female")),
      status_prev = case_when(
        status_prev == 1 ~ "Normal Cognition",
        status_prev == 2 ~ "MCI",
        status_prev == 3 ~ "Dementia",
      ),
      status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")),
      Total_p = as.numeric(Total_p),
      across(c(pred.1:pred.3), as.numeric)) |>
    tidyr::pivot_longer(cols = c(pred.1:pred.3), names_to = "status", values_to = "prob") |>
    mutate(status = case_when(
      status == "pred.1" ~ "Normal Cognition",
      status == "pred.2" ~ "MCI",
      status == "pred.3" ~ "Dementia"
    ),
    status = factor(status, levels = c("Normal Cognition", "MCI", "Dementia")),
    status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")))
}

plot_predictions_stationary <- function(predictions, variable, x_axis) {
  # Visualizes predicted transition probabilities from stationary multi-state models.
  # Shows probability curves by procrastination level, stratified by baseline state.
  # Arguments:
  #   - predictions: Tidy predictions from tidy_predictions()
  # Returns:
  #   - Faceted ggplot showing predicted probability curves
  # 
  predictions |>
    ggplot(aes(x = {{variable}}, y = prob, colour = status_prev)) +
    geom_line(linewidth = 1) +
    ggokabeito::scale_colour_okabe_ito() +
    labs(
      title = "Predicted transition probabilities (stationary model)",
      x = x_axis, y = "Probability", colour = "Previous State") +
    facet_wrap(~ status, labeller = labeller(
      status = function(x){paste0("Transition to: ", x)}
    ))
}

create_additive_predictions <- function(
    data, model, var, var_seq, hold_constant = list(Gender = factor(0), Education_tri = factor(0))) {

  # Set default prediction sequence based on variable type
  if (is.null(var_seq)) {
    if (is.numeric(data[[var]])) {
      predict_seq <- seq(min(data[[var]], max(data[[var]]), length = 200))
    } else if (is.factor(data[[var]])) {
      predict_seq <- levels(data[[var]])
    } else {
      predict_seq <- unique(data[[var]])
    }
  }

  # Create list of all variables needed for prediction
  all_vars <- all.vars(formula(model)[-2])  # Get RHS variables from model formula

  # Create base grid with all variables EXCEPT the prediction variable
  base_grid <- expand.grid(
    lapply(all_vars, function(v) {
      if (v == var) return(NULL)  # Skip prediction variable

      # Special cases - keep all levels
      if (v %in% c("status_prev", "wave")) {
        return(unique(data[[v]]))
      }

      if (v %in% names(hold_constant)) {
        # Use user-specified constant value
        hold_constant[[v]]
      } else if (is.numeric(data[[v]])) {
        # Use mean for numeric variables
        mean(data[[v]], na.rm = TRUE)
      } else if (is.factor(data[[v]])) {
        # Use first level for factors
        factor(levels(data[[v]])[1], levels = levels(data[[v]]))
      } else {
        # Default to most common value
        names(sort(table(data[[v]]), decreasing = TRUE))[1]
      }
    }) |> purrr::compact() |> setNames(all_vars[all_vars != var]),
    KEEP.OUT.ATTRS = FALSE
  )

  # Create full prediction grid
  pred_grid <- base_grid |> 
    tidyr::crossing(!!var := var_seq) |>
    modelr::add_predictions(model = model, var = "pred", type = "probs") |>
    tidy_predictions() |>
    mutate(wave = factor(wave))

  return(pred_grid)
}

plot_additive_predictions <- function(data, var, x_label, subtitle) {

  # Generating predictions
  data |>
    ggplot(aes(x = as.numeric(.data[[var]]), y = prob, color = status_prev)) +
    geom_line(linewidth = 1) +
    ggokabeito::scale_color_okabe_ito() +
    labs(
      title = "Transition Probabilities (Additive Model)",
      subtitle = subtitle,
      x = x_label, 
      y = "Predicted Probability", 
      color = "Previous State"
    ) +
    facet_grid(wave ~ status, labeller = labeller(
      wave = function(x) case_when(
        x == 2 ~ "Year = 2018",
        x == 3 ~ "Year = 2020",
        x == 4 ~ "Year = 2022",
      ),
      status = function(x) paste("Transition to:", x)
    )) +
    theme(
      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
      panel.spacing = unit(1, "lines")) 
}

get_time_varying_matrix <- function(model, time_point) {
  expand.grid(
    Gender = factor(0),
    Age = mean(data_stack$Age),
    Education_tri = factor(0),
    Total_dep_2016 = mean(data_stack$Total_dep_2016),
    Total_p = mean(data_stack$Total_p),
    status_prev = factor(1:3),
    wave = time_point) |> 
    modelr::add_predictions(model, var = "prob", type = "probs") |> 
    select(status_prev, starts_with("prob")) |> 
    mutate(
      status_prev = factor(status_prev, labels = state_names),
      across(starts_with("prob"), ~round(., 3))) |>
    tibble::column_to_rownames("status_prev") |>
    as.matrix() |> `colnames<-`(state_names)
}
```

## Setting theme

```{r set-theme}
#| code-fold: true
#| code-summary: "Check out my theme"

colour <- "#212427"

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, colour = colour, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12, colour = colour),
      axis.title = element_text(size = 10, colour = colour, face = "bold"),
      strip.text = element_text(size = 10, colour = colour, face = "bold"),
      legend.title = element_text(hjust = 0.5, colour = colour, face = "bold"),
      ))
```

## Reading in data {.unnumbered}

```{r data}
data <- read.csv(here::here("analysis/data/data.csv"))
```

@@

```{r covariates}
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
  paste0("Total_dep_", seq(2016, 2022, by = 2)),
  "Total_p")
```

@@

```{r long_data}
data_stack <- data_stack |>
  inner_join(data[, cols], by = "ID") |>
  pivot_and_factorise() |>
  group_by(ID) |>
  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))

@@

```{r model-fitting}
#| code-fold: true
#| code-summary: "Check out my models"
#| 
# Baseline model (only gender; reference: NC)
fit_1a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# Baseline model (only gender; reference: MCI)
fit_1b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# With procrastination (reference: NC)
fit_2a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# With procrastination (reference: MCI)
fit_2b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Full model with previous state (reference NC)
fit_3a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# Full model with previous state (reference MCI)
fit_3b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```

@@

    # Mutating to factors
    transition = factor(
      transition, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term, 
      levels = c("Being female", "Age", "High school degree vs. No education",
                 "Further education vs. No education", "Depression Scores (2016)",
                 "Procrastination (2020)", "Previous state: MCI", 
                 "Previous state: Dementia")),

@@

Let's visualise the estimated odds ratios (@fig-odds-ratio) and the predicted transition probabilities (@fig-model-predictions) for the stationary model .

@@

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-model-predictions
#| fig-cap: "Predicted transition probabilities from the stationary model."
#| 
## Making a prediction dataset -------------------------------------------------
pred_data <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Total_dep_2016 = mean(data_stack$Total_dep_2016),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200))

@@

# Getting estimated probabilities
estimated_probs <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Total_dep_2016 = mean(data_stack$Total_dep_2016),
  Total_p = mean(data_stack$Total_p),
  status_prev = states) |>
  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs")

@@

# Model 4: Additive time effects
fit_4a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev + wave, family = multinomial, 
  data = data_stack, trace = FALSE)

fit_4b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev + wave, family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Model 5: State-specific time effects
fit_5a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + (status_prev * wave), family = multinomial, 
  data = data_stack, trace = FALSE)

fit_5b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + (status_prev * wave), family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Model 6: Full time interactions
fit_6a <- nnet::multinom(
  status ~ (Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev) * wave, family = multinomial, 
  data = data_stack, trace = FALSE)

fit_6b <- nnet::multinom(
  status ~ (Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev) * wave, family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```

>>>>

```{r}
#| label: setup
#| warning: false
#| message: false
#| output: false
#| code-fold: true
#| code-summary: "Check out my code"

# Packages ---------------------------------------------------------------------
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

# Functions --------------------------------------------------------------------
files <- list.files(here::here("R/"), full.names = TRUE)

sapply(files, source)

# Theme ------------------------------------------------------------------------
colour <- "#212427"

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, colour = colour, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12, colour = colour),
      axis.title = element_text(size = 10, colour = colour, face = "bold"),
      strip.text = element_text(size = 10, colour = colour, face = "bold"),
      legend.title = element_text(hjust = 0.5, colour = colour, face = "bold"),
      ))

# Data -------------------------------------------------------------------------
data <- read.csv(here::here("analysis/data/data.csv"))
```

@@

```{r covariates}
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("Total_dep_", seq(2016, 2022, by = 2)),
  "Total_p")
```

@@

```{r long_data}
data_stack <- data_stack |>
  inner_join(data[, cols], by = "ID") |>
  # mutate(Age = Age - (2022 - 2016)) |>  # Age of person in 2016
  # filter(Age >= 50) |>                  # Remove those with were not initially 50+ in 2016
  pivot_and_factorise(time_vary = FALSE) |>
  group_by(ID) |>
  mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) # Make age time varying

@@

```{r model-fitting}
#| code-fold: true
#| code-summary: "Check out my models"
#| 
# Baseline model (only gender; reference: NC)
fit_1a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# Baseline model (only gender; reference: MCI)
fit_1b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# With procrastination (reference: NC)
fit_2a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# With procrastination (reference: MCI)
fit_2b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Full model with previous state (reference NC)
fit_3a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
  family = multinomial, 
  data = data_stack, trace = FALSE)

# Full model with previous state (reference MCI)
fit_3b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
  family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```

@@

    # Mutating to factors
    transition = factor(
      transition, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term, 
      levels = c("Being female", "Age", "High school degree vs. No education",
                 "Further education vs. No education", "Depression",
                 "Procrastination (2020)", "Previous state: MCI", 
                 "Previous state: Dementia")),

@@

Let's visualize the estimated odds ratios (@fig-odds-ratio) and the predicted transition probabilities (@fig-model-predictions) for the stationary model .

@@

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-model-predictions
#| fig-cap: "Predicted transition probabilities from the stationary model."
#| 
## Making a prediction dataset -------------------------------------------------
pred_data <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200))

@@

# Getting estimated probabilities
estimated_probs <- expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Depression = mean(data_stack$Depression),
  Total_p = mean(data_stack$Total_p),
  status_prev = states) |>
  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs")

@@

# Model 4: Additive time effects
fit_4a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev + wave, family = multinomial, 
  data = data_stack, trace = FALSE)

fit_4b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev + wave, family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Model 5: State-specific time effects
fit_5a <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (status_prev * wave), family = multinomial, 
  data = data_stack, trace = FALSE)

fit_5b <- nnet::multinom(
  status ~ Gender + Age + Education_tri + Depression + Total_p + (status_prev * wave), family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)

# Model 6: Full time interactions
fit_6a <- nnet::multinom(
  status ~ (Gender + Age + Education_tri + Depression + Total_p + status_prev) * wave, family = multinomial, 
  data = data_stack, trace = FALSE)

fit_6b <- nnet::multinom(
  status ~ (Gender + Age + Education_tri + Depression + Total_p + status_prev) * wave, family = multinomial, 
  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
  trace = FALSE)
```
